\chapter{Hello, \texorpdfstring{\LaTeX}{LaTeX}!}

Now that you have a \LaTeX{} distribution successfully installed,
let's try it out.
Open up your editor of choice and save the following as \texttt{hello.tex}:
\begin{leftfigure}
\begin{lstlisting}
\documentclass{article}
% Say hello
\begin{document}
Hello, World!
\end{document}
\end{lstlisting}
\end{leftfigure}
We then want to run this through \LaTeX{} (the program)\footnote{Not to be
confused with \LaTeX{} the lunchbox, \LaTeX{} the breakfast cereal,
or \LaTeX{} the flamethrower. The kids love this stuff!}
to get our document.
The installation will have placed several different versions---or
\introduce{engines}---on your machine,
but for the entire book, we'll always use \LuaLaTeX or \XeLaTeX.
These are the newest engines available---see Appendix~\ref{history} for an
explanation of how they differ from the others.

If you are using a \LaTeX{}-specific editor, it may contain a drop-down menu
or some other configuration to select the engine you'd like to use,
as well as a button to generate your document.
Otherwise, from a terminal,\punckern\footnote{How to work a terminal emulator,
making sure the newly-installed \LaTeX{} programs are in your \texttt{PATH},
and so on are all outside the scope of this book.
As is tradition, the leading dollar signs in examples just denote a console
prompt, and shouldn't actually be typed.}
run the following:
\begin{leftfigure}
\begin{lstlisting}
$ xelatex hello.tex
\end{lstlisting}
\end{leftfigure}
Feel free to use \texttt{lualatex} instead---there are a few differences
between the two, but either is fine for now.
With luck, you should see some output that ends in a message like:
\begin{leftfigure}
\begin{lstlisting}
Output written on hello.pdf (1 page).
Transcript written on hello.log.
\end{lstlisting}
\end{leftfigure}
and in your current directory, a newly minted \texttt{hello.pdf}.
Open it up and you should find a page with this at the top:
\begin{leftfigure}
\lm Hello, World!
\end{leftfigure}

Congrats!
You just created your first document.
Let's unpack what we did.
All \LaTeX{} documents begin with a \verb|\documentclass| declaration,
which picks a style to use.
Many classes are available---and you can even create your own---but common
ones include \texttt{article}, \texttt{report}, \texttt{book},
and \texttt{beamer}.\punckern\footnote{This last one is for slideshows.
Kind of an odd name, no?}
For your average document, \texttt{article} is probably a good choice.
Next, is \verb|% Say hello|.
This is a \introduce{comment}---anything placed after a percent symbol on a line
is ignored by the engine,
so we can use \verb|%| to leave notes about what's going on to anybody reading
the document's source.\punckern\footnote{Including, perhaps most importantly,
a confused version of your future self!}
Finally, we use \verb|\begin{document}| to tell \LaTeX{} that what follows
is our actual contents,
and we use \verb|\end{document}| to state that we are finished.

So far, so good.
Let's cover some more basics.

\section{Spacing}

\LaTeX{} generally handles inter-word spacing for you, regardless of how many
spaces\footnote{Or tabs} you type. For example,
\begin{leftfigure}
\begin{lstlisting}
The number  of   spaces    between words doesn't   matter. The same
is true for sentences.

An empty line starts a new paragraph.
\end{lstlisting}
\end{leftfigure}
yields
\begin{leftfigure}
\lm The number  of   spaces    between words doesn't   matter. The same
is true for sentences.

An empty line  starts a new paragraph.
\end{leftfigure}
Notice that \LaTeX{} is automatically following typographic
conventions, such as indenting new paragraphs and  leaving more space after a
period than one does between words.
One quirk to be aware of is that comments ``eat'' all of the leading
space on the subsequent line, such that
\begin{leftfigure}
\begin{lstlisting}
This% weird, right?
  is strange
\end{lstlisting}
\end{leftfigure}
gives
\begin{leftfigure}
This% weird, right?
  is strange
\end{leftfigure}

\section{Commands}

\LaTeX{} has various commands to issue instructions to
the engine, and you can define your own as well.
Their names always begin with a backslash (\,\texttt{\textbackslash}\,),
contain only letters, and are case-sensitive.\punckern\footnote{%
% \verb doesn't play nicely in footnotes, so...
\texttt{\textbackslash foo}
is different from \texttt{\textbackslash Foo}, for example.}
Some commands require parameters, or \introduce{arguments}---\verb|\documentclass|,
for example, needs to know which document class we want.
These arguments are enclosed in subsequent pairs of braces,
so if some command needed two arguments, we would type:
\begin{leftfigure}
\begin{lstlisting}
\somecommand{argument1}{argument2}
\end{lstlisting}
\end{leftfigure}
Some commands also take optional arguments, which are enclosed in brackets
and precede the mandatory ones.
Say you want to inform \LaTeX{} that your document will be printed as
double-sided pages.\punckern\footnote{\texttt{twoside} introduces commands
that only make sense in the context of double-sided printing,
such as ones that skip to the start of the next odd page.
It also allows you to have different margins for even and odd pages,
which is useful for things like this book.}
This is done with an optional argument to \verb|\documentclass|:
\begin{leftfigure}
\begin{lstlisting}
\documentclass[twoside]{article}
\end{lstlisting}
\end{leftfigure}

Some commands take no arguments at all, such as \verb|\LaTeX|,
the one that prints the \LaTeX{} logo.
Know that these commands consume any space that follows them.
For example,
\begin{leftfigure}
\begin{lstlisting}
\LaTeX is great, but it can be strange sometimes.
\end{lstlisting}
\end{leftfigure}
will give you
\begin{leftfigure}
\lm \LaTeX is great, but it can be strange sometimes.
\end{leftfigure}
You can fix this by adding an empty pair of braces to the command.
Of course, the braces aren't needed if there is no space to preserve:
\begin{leftfigure}
\begin{lstlisting}
Let's learn \LaTeX! \LaTeX{} is a powerful tool,
but a few of its rules are a little weird.
\end{lstlisting}
\end{leftfigure}
gets us
\begin{leftfigure}
\lm Let's learn \LaTeX! \LaTeX{} is a powerful tool,
but a few of its rules are a little weird.
\end{leftfigure}

\section{Special characters and line breaks}

Some characters have special meanings in \LaTeX.
We saw above, for example, that \verb|%| starts a comment
and \verb|\| marks the start of some command.
The full list is:
\begin{leftfigure}
\begin{lstlisting}
# $ % ^ & _ { } ~ \
\end{lstlisting}
\end{leftfigure}
All have corresponding commands
if you want to make them appear in your document.
\begin{leftfigure}
\begin{lstlisting}
\# \$ \% \^{} \& \_ \{ \} \~{} \textbackslash
\end{lstlisting}
\end{leftfigure}
will produce
\begin{leftfigure}
\lm \# \$ \% \^{} \& \_ \{ \} \~{} \textbackslash
\end{leftfigure}
Regardless of what comes after them, you must always add braces to
the caret (\,\texttt{\^{}}\,) and tilde (\,\~{}\,).
This is a relic from days when these characters were used to produce
\introduce{diacritical marks}:
once upon a time, users had to typeset ``jalapeño'' as
\verb|jalape\~no|.
Today, we just type \texttt{ñ} into our source
file.\punckern\footnote{The ease with which you can do this depends
on the keyboard you're using, the language settings in your \acronym{os},
and your editor.
Later on, we'll talk much more about non-English languages and Unicode fun.}

If you're wondering why we print \texttt{\textbackslash} with
\verb|\textbackslash| instead of just \verb|\\|,
it's because the latter is the command to force a line break.
\begin{leftfigure}
\begin{lstlisting}
Give me \\
a brand new line!
\end{lstlisting}
\end{leftfigure}
obeys:
\begin{leftfigure}
\lm Give me \\
a brand new line!
\end{leftfigure}
Use this power judiciously---deciding how to break paragraphs into lines
automatically is one of \LaTeX{}'s best skills.

\section{Environments}

We often control our document by organizing text into \introduce{environments}.
Environments always start with \verb|\begin{name}| and end with \verb|\end{name}|,
where \texttt{name} is the name of the environment.
All documents use the \texttt{document} environment to start and end their bodies,
but there are many others.
Take \texttt{quote}, which adds additional spacing on both sides of a block
quotation:
\begin{leftfigure}
\begin{lstlisting}
Donald Knuth once wrote,
\begin{quote}
We should forget about small efficiencies,
say about 97\% of the time:
premature optimization is the root of all evil.
Yet we should not pass up our opportunities in that critical 3\%.
\end{quote}
\end{lstlisting}
\end{leftfigure}
gives us:
\begin{leftfigure}
\lm
Donald Knuth once wrote,
\begin{quote}
We should forget about small efficiencies,
say about 97\% of the time:
premature optimization is the root of all evil.
Yet we should not pass up our opportunities in that critical 3\%.
\end{quote}
\end{leftfigure}

\section{Groups and command scope}
Some commands change how \LaTeX{} typesets text.
\verb|\itshape|, for example, \textit{italicizes} everything that comes after it.
If we want to limit a command's effect to a certain region, we can surround it
with braces.
\begin{leftfigure}
\begin{lstlisting}
{\itshape Sometimes we want italics}, but only sometimes.
\end{lstlisting}
\end{leftfigure}
is set as
\begin{leftfigure}
\lm {\itshape Sometimes we want italics}, but only sometimes.
\end{leftfigure}
The text within the braces forms a \introduce{group},
and all commands issued inside the group only take effect until it ends.
Environments also also form implicit groups:
\begin{leftfigure}
\begin{lstlisting}
\begin{quote}
\itshape If I italicize a quote, the following text will
use upright type again.
\end{quote}
See? Back to normal.
\end{lstlisting}
\end{leftfigure}
produces
\begin{leftfigure}
\lm
\begin{quote}
\itshape If I italicize a quote, the following text will
use upright type again.
\end{quote}
See? Back to normal.
\end{leftfigure}

One final use of groups is to get around the spacing oddities of zero-argument
commands: some prefer \verb|{\LaTeX}| over \verb|\LaTeX{}|.
